---
title: "Iris Dataset"
author: "David C. King"
date: "2/17/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
bibliography: iris.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE # do not interrupt in case of errors
                      )

# install packages if needed
if (! require(reshape2)) {install.packages("reshape2")}
if (! require(ggplot2)) {install.packages("ggplot2")}

library(reshape2)
library(ggplot2)
```


# Builtin datasets
R has a package called "datasets" that provides many old but readily available datasets to test things out on.

In the console, do `?iris`

The help window will gives a description of the data, and the format of the `iris` and `iris3` data frames.

To see the full list of datasets in the package, click on 'Index' at the bottom of the help page.

# `iris` dataset

### Description

This famous (Fisher's [@Fisher1936] or Anderson's [@Anderson1936]) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris *setosa*, *versicolor*, and *virginica.*


```{r iris dataframe}

data(iris) # copies the data frame into your workspace
head(iris) # see the first 10 rows. A data frame is basically a spreadsheet
iris$Petal.Length # access a whole column by name using the '$'
iris[[3]] # or by its column number in [[ ]] 
```
# Visualizing the iris data

How do we get a sense of Sepal and Petal dimensions among iris species? Given these measurements, let's look at some of their properties.

## Boxplots

The distributions of each measurement reveal a greater variability in certain species.

```{r individual boxplots, include=FALSE}
ggplot(iris, aes(x=Sepal.Length, col=Species)) + geom_boxplot() 
ggplot(iris, aes(x=Sepal.Width, col=Species)) + geom_boxplot() 
ggplot(iris, aes(x=Petal.Length, col=Species)) + geom_boxplot() 
ggplot(iris, aes(x=Petal.Width, col=Species)) + geom_boxplot() 
```

To make a paneled version of the four boxplots, we make a new data structure. I.e., stick all the value columns together vertically, then add a descriptive column as to which measurement they are.

```{r wide versus long data frame}
library(reshape2) # for melt
iris_long = melt(iris, 
               id = "Species",
               value.name = "centimeters", 
               variable.name = "flower_part") 

head(iris_long) 
```

### Split into different boxplots based on the flower part.

```{r boxplots with panels face_wrap}
ggplot(iris_long, aes(x = centimeters, col = Species)) + 
  geom_boxplot() + 
  facet_wrap( ~ flower_part)
```

The most variable measurements appear to be the sepal and petal lengths of *versicolor* and *virginica*. The least amount of variation appears to be in the petal width and lengths of *setosa*.

*setosa* is easiest to distinguish from the other species.

### Split the boxplots by species instead of flower part.
```{r }
ggplot(iris_long, aes(x=centimeters, col=flower_part)) + 
  geom_boxplot() + 
  facet_wrap(~Species) 
```

Again, *setosa* has the most characteristic pattern, whereas *versicolor* and *virginica* are more similar to each other.


## Scatter plots

Comparing measurements is a way to get a sense of the data. Plot the columns in the data frame against each other.

```{r traditional iris plots, include=FALSE}
# plot two columns against each other
plot(iris$Sepal.Length,iris$Sepal.Width, main="main title")
summary(iris)

# add meaningful labels
plot(iris$Sepal.Length,iris$Sepal.Width, main="Sepal length vs width", xlab="Sepal Length", ylab="Sepal Width")

plot(iris$Sepal.Length,iris$Sepal.Width, main="Sepal length vs width", xlab="Sepal Length", ylab="Sepal Width", col=iris$Species)

# you can change the color with "col"
plot(iris$Sepal.Length,iris$Sepal.Width, main="Sepal length vs width", xlab="Sepal Length", ylab="Sepal Width", col="purple") # col can be a "named" color 

# There are about 650 of them. Let's look at the ends of the list
head(colors())
tail(colors())

# color can be a number
plot(iris$Sepal.Length,iris$Sepal.Width, main="Sepal length vs width", xlab="Sepal Length", ylab="Sepal Width", col=2)

# factors can be used as numbers
class(iris$Species)
plot(iris$Sepal.Length,iris$Sepal.Width, main="Sepal length vs width", xlab="Sepal Length", ylab="Sepal Width", col=iris$Species)

legend(7,4.5, legend=levels(iris$Species),col=seq_along(levels(iris$Species)),pch=21)

# factors are a special type that can be used as a "grouping" variable
# R uses them as numbers
levels(iris$Species)

```

```{r iris-ggplots, echo=FALSE}
library(ggplot2)
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width,col=Species)) + 
  geom_point() + 
  ggtitle("Iris Dataset", subtitle = "Sepal Length x Width")

ggplot(iris, aes(x=Sepal.Length, y=Petal.Length,col=Species)) + 
  geom_point() + 
  ggtitle("Iris Dataset", subtitle = "Sepal Length x Petal Length")

ggplot(iris, aes(x=Sepal.Length, y=Petal.Width,col=Species)) + 
  geom_point() + 
  ggtitle("Iris Dataset", subtitle = "Sepal Length x Petal Width")

```
Different sepal to petal comparisons seem to separate better than others.

### Adding a trendline with `stat_smooth()`

```{r scatterplot with trendline}
ggplot(iris, aes(x=Sepal.Length, y=Petal.Length,col=Species)) + 
  geom_point() + 
  ggtitle("Iris Dataset", subtitle = "Sepal Length x Petal Length") + 
  geom_point(alpha = 0.3,  position = position_jitter()) + 
  stat_smooth(method = "lm")
```

It seems that petal length is tightly distributed when compared to sepal length. Perhaps these relationships could be combined to help distinguish these species?

## Conclusion

A combination of sepal and petal measurements may elicit the best way of distinguishing *versicolor* from *virginica*, whereas *setosa* is quite distinct in these properties.

## Bibliography